Mapping short reads, faithfully

### 2020-02-10 manuscript.tex
Merge benchmark and implementation details in results

The results reveal two essential insights: The first is that the
asymptotic decay of the failure rate is approximately constant when the
read maps to a location with 20 or fewer duplicates. Beyond this, the
asymptotic decay slows down and the recall decreases markedly. The second
insight is that up to 100 nucleotides, the failure rate is always above 1
error per 100,000 reads. In other words, there is no hope to map a read
with high confidence using MEM seeds if the target has a paralog.

These insights suggest that an efficient strategy is to use a fast mapping
mode when the target has many duplicates because the read cannot be mapped
with high confidence anyway. In contrast, when the number of duplicates is
low, it is important to know whether the target is a unique sequence
because this is the only way to map a read with high confidence. So after
MEM seeding, our strategy is to quickly estimate the number of duplicates,
find a decent candidate location as fast as possible and finally estimate
how reliable this location is. In other words, we search the reads that
can be mapped with high confidence instead of searching a high-confidence
location for every read. A high-level summary of this strategy is shown in
Figure~\ref{fig_tree}.

@@

The key in the flow chart of Figure~\ref{fig_tree} is to dispatch the
reads based the number $N$ of paralogs of the target, which is an
indirect measure of the mappability of the read~\cite{pmid22276185}.
Estimating $N$ must be fast so that this step does not become a
bottleneck. To achieve this, we propose to use the seeding process itself.
MEM seeds are best computed using the backward
search~\cite{ferragina2000opportunistic}, an algorithm that returns the
number of hits in the genome as the query is extended backward. If the
target has no or few paralogs (Figure~\ref{fig_back}a), the number of hits
will quickly go below 2 (when there is only one hit, it is most likely the
target itself). In contrast, if the target has many paralogs in the genome
(Figure~\ref{fig_back}b), the number of hits will remain at 2 or above for
may iterations.

@@

We designed another estimation procedure based on the backward search,
where this time we use the best candidate location and not the read as a
query. The principle, sketched in Figure~\ref{fig_supertest}a, is to
perform the backward search on 20-mers and to test whether the genome
contains a single hit (the candidate location). The rationale is that the
total number of 20-mers is approximately $10^{12}$, so the chances of
spurious random hits are very low, even in the largest known genomes.

@@

The extremely high mapping quality of super reads comes from an unexpected
property of the test depicted in Figure~\ref{fig_supertest}. To see why
they can be mapped with such confidence, consider a read aligning
perfectly with an incorrect sequence of the genome. Observe that for
this to happen, the read must contain an error that is perfectly
compensated in some duplicate of the target. This can only happen if $i.$
the target has a duplicate, $ii.$ the read contains an error, $iii.$ this
error matches the duplicate and $iv.$ the target and the duplicate are
otherwise identical.

@@

We implemented the mapping strategy presented in Figure~\ref{fig_tree} in
a prototype mapper called MEM Mapper Prototype (MMP). We used plain
mapping algorithms to better highlight the benefits of building a mapper
for faithfulness. The index consists of a standard FM-index of the genome
and its reverse complement~\citep[the implementation is detailed in
ref.][]{blog}. We also added an auxiliary lookup table storing the states
of the bacwkard search for all possible 12-mers, allowing us to skip the
first 12 iterations for every query. This lookup table has a fixed size of
256~MB. Sequence alignment is carried out using a version of the
Needleman-Wunsch algorithm~\cite{pmid5420325} with an option to abort if
the score of the current best hit is exceeded (continuing the algorithm is
useless to find the optimal location of the read). Mapping quality is
computed as detailed in the Materials and Methods section.

We set up a test using the genomes of \textit{Drosophila melanogaster}
(the fruit fly), \textit{homo sapiens} (modern humans) and \textit{Pinus
taeda} (a north American species of pine). The high-level features of the
genomes are summarized in Table~\ref{table_genomes}. We included
\textit{Drosophila} and human as model organisms with high-quality genome
assemblies, and pine as a non-model organism with a very complex genome
and a draft assembly.

@@

\subsection{MMP implementation}

MMP is implemented as a stand-alone C program including the code of
\texttt{sesame}~\cite{Filion619155} and of
\texttt{divsufsort}~\cite{fischer2017dismantling}. MMP uses a standard
implementation of the FM-index for the genome and its reverse complement,
as detailed in~\cite{blog}, and of the Needleman-Wunsch
algorithm~\cite{pmid5420325}, with the modification that the algorithm is
interrupted if the score rises above that of the current best hit. During
indexing, the backward search~\cite{ferragina2000opportunistic} is carried
out on all possible 12-mers and the results are stored so that the first
12 iterations of every query can be skipped during the mapping.

The implementation otherwise follows the flow chart of
Figure~\ref{fig_tree} using the formula above, with two differences.
The first is that super reads can have 0, 1 or two mismatches, as long as
all the extracted 20-mers have a single hit in the genome. How mapping
quality is computed in those cases is described in the Supplementary
Material. The second difference is that once the best hit is found, $N$ is
estimated again on the genome instead of the read, and the higher of the
two estimates is kept.

@@

Here we designed and implemented a strategy to map short reads faithfully.
The principles are based on key insight gained from models to compute
seeding probabilities~\cite{Filion619155}, showing that MEM seeds are
efficient only when the target has few paralogues (Figure~\ref{fig_MEM}).
This motivates the design of a strategy where we first estimate the copy
number of the target (Figure~\ref{fig_back}) to dispatch reads to
different mapping subroutines (Figure~\ref{fig_tree}). This amounts to
allocating more resources to the most mappable reads~\cite{pmid22276185}.
This process revealed the existence of super reads, \textit{i.e.}, reads
that can be mapped with extremely high confidence
(Figures~\ref{fig_supertest} and \ref{fig_supread}), together setting the
basis of an efficient mapping strategy. We wrote a mapper based on this
strategy and we showed that it is competitive with BWA-MEM and Bowtie
(Figures~\ref{fig_faith}, \ref{fig_acc} and \ref{fig_timem}). Importantly,
the mapping process itself relies only on standard algorithm, so all the
benefits of the mapper come from faithfulness. Overall, this demonstrates
that improving faithfulness is a fruitful strategy to improve short-read
mappers.

>>>>

The results reveal two essential insights: The first is that the
asymptotic decay of the failure rate is approximately constant when the
read maps to a location with 20 or fewer duplicates. Beyond this, the
asymptotic decay slows down and the recall decreases markedly. The second
insight is that up to 100 nucleotides, the failure rate is always above 1
error per 100,000 reads. In other words, there is no hope to map a read
with very high confidence using MEM seeds if the target has a paralog.

These insights suggest that an efficient strategy is to use a fast mapping
mode when the target has many duplicates because the read cannot be mapped
with high confidence anyway. In contrast, when the number of duplicates is
low, it is important to know whether the target is a unique sequence
because this is the only way to map a read with high confidence. So after
MEM seeding, our strategy is to quickly estimate the number of duplicates,
find a decent candidate location as fast as possible and finally estimate
how reliable this location is. In essence, we search the reads that
can be mapped with high confidence instead of searching a high-confidence
location for every read. A high-level summary of this strategy is shown in
Figure~\ref{fig_tree}.

@@

The key in the flow chart of Figure~\ref{fig_tree} is to dispatch the
reads based the number $N$ of paralogs of the target, which is an
indirect measure of the mappability of the read~\cite{pmid22276185}.
Estimating $N$ must be fast so that this step does not become a
bottleneck. To achieve this, we propose to use the seeding process itself.
MEM seeds are best computed using the backward
search~\cite{ferragina2000opportunistic}, an algorithm that returns the
number of hits in the genome as the query is extended backward. If the
target has no or few paralogs (Figure~\ref{fig_back}a), the number of hits
will quickly go below 2 (when there is only one hit, it is most likely the
target itself). In contrast, if the target has many paralogs in the genome
(Figure~\ref{fig_back}b), the number of hits will remain at 2 or above for
many additional iterations.

@@

We designed another estimation procedure based on the backward search,
where this time we use the best candidate location and not the read as a
query. The principle, sketched in Figure~\ref{fig_supertest}a, is to
perform the backward search on 20-mers and to test whether the genome
contains a single hit. The rationale is that the
total number of 20-mers is approximately $10^{12}$, so the chances of
spurious random hits are very low, even in the largest known genomes.

@@

The extremely high mapping quality of super reads comes from an unexpected
property of the test depicted in Figure~\ref{fig_supertest}. To see why
they can be mapped with such confidence, consider a read aligning
perfectly to an incorrect sequence of the genome. Observe that for
this to happen, the read must contain one or more errors that are perfectly
compensated in some duplicate of the target. This can only happen if $i.$
the target has a duplicate, $ii.$ the read contains an error, $iii.$ this
error matches the duplicate and $iv.$ the target and the duplicate are
otherwise identical.

@@

We implemented the mapping strategy presented in Figure~\ref{fig_tree} in
a prototype mapper called MEM Mapper Prototype (MMP). MMP is implemented
as a stand-alone C program including the code of
\texttt{sesame}~\cite{Filion619155} and of
\texttt{divsufsort}~\cite{fischer2017dismantling}. We used plain
mapping algorithms to better highlight the benefits of building a mapper
for faithfulness. The index consists of a standard FM-index of the genome
and its reverse complement~\citep[the implementation is detailed in
ref.][]{blog}. We also added an auxiliary lookup table storing the states
of the bacwkard search for all possible 12-mers, allowing us to skip the
first 12 iterations for every query. This lookup table has a fixed size of
256~MB. Sequence alignment is carried out using a version of the
Needleman-Wunsch algorithm~\cite{pmid5420325} with an option to abort if
the score of the current best hit is exceeded (continuing the algorithm is
useless to find the optimal location of the read). Mapping quality is
computed as detailed in the Materials and Methods section.

The implementation otherwise follows the flow chart of
Figure~\ref{fig_tree} using the formulae above, with two differences.
The first is that super reads can have 0, 1 or two mismatches, as long as
all the extracted 20-mers have a single hit in the genome. How mapping
quality is computed in those cases is described in the Supplementary
Material. The second difference is that once the best hit is found, $N$ is
estimated again using the genomic sequence instead of the read, and the
higher of the two estimates is kept.

\subsubsection{Benchmark results}
We set up a test using the genomes of \textit{Drosophila melanogaster}
(the fruit fly), \textit{homo sapiens} (modern humans) and \textit{Pinus
taeda} (a north American species of pine). The high-level features of the
genomes are summarized in Table~\ref{table_genomes}. We included
\textit{Drosophila} and human as model organisms with high-quality genome
assemblies, and pine as a non-model organism with a very complex genome
and a draft assembly.

@@

@@

Here we designed and implemented a strategy to map short reads faithfully.
The principles are based on key insight gained from models to compute
seeding probabilities~\cite{Filion619155}, showing that MEM seeds are
efficient only when the target has few paralogs (Figure~\ref{fig_MEM}).
This motivates the design of a strategy where we first estimate the copy
number of the target (Figure~\ref{fig_back}) to dispatch reads to
different mapping subroutines (Figure~\ref{fig_tree}). This amounts to
allocating more resources to the most mappable reads~\cite{pmid22276185}.
This process revealed the existence of super reads, \textit{i.e.}, reads
that can be mapped with extremely high confidence
(Figures~\ref{fig_supertest} and \ref{fig_supread}), together setting the
basis of an efficient mapping strategy. We wrote a mapper based on this
strategy and we showed that it is competitive with BWA-MEM and Bowtie
(Figures~\ref{fig_faith}, \ref{fig_acc} and \ref{fig_timem}). Importantly,
the mapping process itself relies only on standard algorithm, so all the
benefits of the mapper come from faithfulness. Overall, this demonstrates
that improving faithfulness is a fruitful strategy to improve short-read
mappers.
